Annotations on Quantum Pseudorandomness and $k$-Designs

### 2020-11-23 PHYS790/final.tex
Added more figures for comparision theorem.

The power of randomness in computation is a cornerstone of contemporary computation theory as evidenced by decades of beautiful, celebrated results. Quantum Information has, by no means, become stranger to this trend. Numerous applications of random unitaries have been discovered in the areas of quantum tomography, quantum cryptography, and, rather naturally, quantum algorithms. To gain access to uniformly-sampled elements from $\mathbb{U}(d)$, one must sample from the Haar distribution over $\mathbb{U}(d)$ where $\mathbb{U}(d)$ denotes the group of unitary $d \times d$ matrices. One method of sampling from the Haar distribution is to apply local random unitaries sampled from distributions over lower-dimensional unitary groups. For example, if we consider an $n$-qubit system, one method of sampling from the Haar distribution over $\mathbb{U}(2^n)$ would be through the following scheme:
\begin{enumerate}
  \item Sample from the Haar distribution over $\mathbb{U}(4)$ to produce a random gate $G$
  \item Uniformly pick two distinct qubits and locally apply $G$ to the pair.
\end{enumerate}
However, this becomes computationally infeasible as even reasonably approximating the Haar distribution under this manner requires an exponential number of local 2-qubit gates \cite{knill}. This hurdle motivates us to study the construction of efficient $k$-designs i.e distributions over $\mathbb{U}(4)$ such that sampling over these distributions and applying our scheme above will converge to good approximations of the $k^{th}$ moments of the Haar distributon over $\mathbb{U}(2^n)$. To this end, we begin with annotations of the seminal paper by Harrow and Low \cite{harrlow}. We first introduce concepts germane to the area of efficient $k$-designs and summarize the techniques employed in the analyses presented in the paper.

@@

\subsection{Definitions and Motivations}
In \cite{harrlow}, the authors tackle the problem of approximate $2$-designs, specifically \emph{unitary} $2$-designs. The central result of the paper claims that the scheme of applying sampled $2$-qubit unitaries from certain distributions with a $k$-gapped property converge to the second moment of the Haar distribution rapidly. The crux of the rapid convergence argument relies on mapping the evolution of an initial state under our random circuit to that of a Markov chain's.
Consequently, the analysis becomes amenable to classic techniques created to study mixing times of Markov chains. To begin, we first state the relevant definitions as introduced by Harrow and Low.
%
\begin{remark}
  As dissecting every proof in \cite{harrlow} may unnecessarily length this report, we occasionally omit proofs of lemmas. Some major theorems will have their proofs delegated to the appendix of this report.
\end{remark}

@@

As discussed in the introduction above, the unitary constructed by sampling from a universal gate set and applying them to random pairs of qubits will converge to a unitary sampled from Haar distribution. However, we would like to find \emph{efficient approximate} $k$-designs. Here, efficient refers to random circuits of length polynomial in the number of qubits. The authors actually prove some stronger, namely that \emph{$k$-copy gapped} distributions suffice.

@@

  then we say that $\mathcal{G}_\mu$ is $k$-copy gapped if $\mathcal{G}_\mu$ has only $k!$ unit eigenvalues.

@@

\noindent A more general definition of universality of gatesets can be defined as well:

@@

  Let $\mu$ be a distribution over $\mathbb{U}(d)$. $\mu$ is deemed to be \emph{universal} for $\mathbb{U}(d)$ if for any open ball $B \subset \mathbb{U}(d)$, there exists a sufficiently large integer $\ell$ such that $\mu^{\star \ell}(B) > 0$

@@

  This is the convolution product in the context of \emph{measures}. For example, if we let $\mu,\sigma$ be two finite Borel measures over $\mathbb{R}$, then we have a new finite Borel measure on $\mathbb{R}$, $\mu \star \sigma$:
  $$ \mu \star \sigma(A) = \int_{\mathbb{R}}\int_{\mathbb{R}} \mathbb{I}_A(x+y) d\mu(x)d\sigma(y)$$
  where $\mathbb{I}$ is the characteristic function for measurable set $A \subset \mathbb{R}$.

@@

\noindent Intuitively, universality implies that sampling enough times over $\mu$ is enough to approximate any gate in $\mathbb{U}(d)$. This aligns with the definition introduced for discrete measures $\mu$ over finite gate sets. We will prove later that any universal gate set will be $k$-copy gapped on $\mathbb{U}(4)$ (Lemma \ref{universal}). For now, let us state the main theorem and its auxiliary lemma to guide our upcoming analysis:

@@

  Let $\rho$ be an initial state of an $n$-qubit system i.e $\rho \in (\mathbb{C}^{2})^{\otimes n}$. We can expand $\rho$ in terms of the Pauli basis $\{\sigma_i\}_{0 \leq i \leq 3}$ as follows:
    \rho = \frac{1}{2^n} \sum_{p_1,p_2} \gamma_0(p_1,p_2) \sigma_{p_1} \otimes \sigma_{p_2}, \quad \gamma_0(p_1,p_2) \in \in \mathbb{C}
  where $p_1,p_2$ run over all strings of Pauli indices of length $n$, $\{0,1,2,3\}^n$. Let $\gamma_W(p_1,p_2)$ be the coefficient of the term $\sigma_{p_1} \otimes \sigma_{p_2}$ for the output state after applying random circuit $W$, $\rho_W = W^{\otimes 2} \rho (W^\dagger)^{\otimes 2}$. Extending this notation allows us to express the expected coefficient after $t$-samples from our distribution $\mu$ as
  $\gamma_t(p_1,p_2) = \mathbb{E}_W [\gamma_W(p_1,p_2)]$. Note that the expectation is taken over all $t$-length random circuits.
\end{notation}

\begin{lemma} \label{mainlemma}
  Let $\rho$ be a an initial state such that $\gamma_0(p,p) \geq 0$ and $\sum_{p} \gamma_0(p,p) = 1$ with $\mu$ and $W$ as in Theorem \ref{maintheorem}. Then there exists constant $C$ such that for $\epsilon > 0$

@@

  \begin{equation}
    \sum_{p_1,p_2 \neq 00} \left| \gamma_t(p_1,p_2) - \delta_{p_1p_2} \frac{1}{2^n(2^n + 1)} \right| \leq \epsilon
  \end{equation}
  for $t \geq n(n + \log 1/\epsilon)$
\end{lemma}

\begin{corollary} \label{maincorollary}
  Let $\mu$, $W$, $\gamma_W$ be as above. Then for \emph{any} initial state $\rho$, there exists constant $C$ such that for $\epsilon > 0$,
  \begin{align}
    & \sum_{p_1,p_2 \neq 00} \left( \gamma_t(p_1,p_2) - \delta_{p_1p_2}(\frac{\sum_{p\neq 0} \gamma_0(p,p)}{4^n - 1} \right)^2 \leq \epsilon \\
    & \sum_{p_1,p_2 \neq 00} \left| \gamma_t(p_1,p_2) - \delta_{p_1p_2}(\frac{\sum_{p\neq 0} \gamma_0(p,p)}{4^n - 1} \right| \leq \epsilon
  \end{align}
  both for $t \geq Cn(n + \log 1/ \epsilon)$
\end{corollary}

@@

We turn back to the general case and begin to define concepts and definitions for group $\mathbb{U}(d)$ and for $d$-qudit states. Set a Hermitian basis of $d \times d$ to be $\sigma_0,\sigma_1,\cdots \sigma_{d^2 -1}$ such that $\sigma_0 = I$. These basis elements span the $d$-qudit states. Here we stress that we sample our unitaries over the \emph{Haar distribution} over $\mathbb{U}(d)$. We denote sampling from the Haar distribution with the notation $U \sim \mathbb{U}(d)$ when necessary. Furthermore, our Hermitian basis operators are such that $Tr(\sigma_p\sigma_q) = d \cdot \delta_{p,q}$ i,e orthogonal under the Hilbert-Schmidt inner product. \newline

\noindent Let ${\bf p} = (p_1, \cdots, p_k) \in \{0,1,\cdots, d^2 -1\}^k$ denote a $k$-string of basis indices. We will be interested in the quantity:
%
\begin{equation} \label{Tdef}
  T({\bf p}) := \mathbb{E}_{U \sim \mathbb{U}(d)} \left[ U^{\otimes k} \sigma_{p_1} \otimes \sigma_{p_2} \otimes \cdots \otimes \sigma_{p_k} (U^\dagger)^{\otimes k} \right]
\end{equation}
This will be the expected state after applying a sampled unitary $U$ from the uniform distribution over $\mathbb{U}(d)$. Naturally, we can reexpress this resulting state into the $\sigma_p$ basis:
%
\begin{equation}
  T({\bf p}) = \sum_{{\bf q}} \widehat{G}(\textbf{p},\textbf{q}) \sigma_{q_1} \otimes \cdots \otimes \sigma_{q_k}
\end{equation}
where the same is taken over all $k$-strings in $\{0, \cdots, d^2-1\}^k$. This allows us to define the matrix $\widehat{G}$ where the rows and columns are indiced by $k$-strings with entries $\widehat{G}(\textbf{q},\textbf{p})$. This matrix turns out to be a \emph{symmetric projector}.

\begin{lemma}
  The matrix $\widehat{G}$ above is symmetric i.e. $\widehat{G}(\textbf{p},\textbf{q}) = \widehat{G}(\textbf{q},\textbf{p})$
\end{lemma}
%
\begin{proof}
  By orthogonality:
  \begin{align}
    \widehat{G}(\textbf{q},\textbf{p})
    & = d^{-k} \cdot Tr(\sigma_{q_1} \otimes \cdots \otimes \sigma_{q_k} T(\textbf{p}) ) \\
    & = d^{-k} \mathbb{E}_{U}\left[ Tr(\sigma_{q_1} \otimes \cdots \otimes \sigma_{q_k}) U^{\otimes k} (\sigma_{p_1} \otimes \sigma_{p_2} \otimes \cdots \otimes \sigma_{p_k}) (U^\dagger)^{\otimes k}\right]
  \end{align}
\end{proof}
%
Note that the expectation and the trace operator commute by linearity of latter. The result follows from the cyclic invariance of trace.
%
\begin{lemma} \label{permfix}
$\widehat{G}$ fixes permutation operators $P_{\pi}, \; \pi \in S_k$.
\end{lemma}
%
These two results are all that is required to show that $\widehat{G}$ is indeed a projector.
%
\begin{theorem} \label{g-proj}
  $\widehat{G}$ is a projector i.e. $\widehat{G}^2 = \widehat{G}$
\end{theorem}
%
\begin{proof}
  Expanding the matrix product shows that
  \begin{align}
     \widehat{G}(\textbf{p},\textbf{q})
     & = \sum_{\textbf{q'}} \widehat{G}(\textbf{p},\textbf{q'})\widehat{G}(\textbf{q'},\textbf{q}) = \\
     & = \sum_{\textbf{q'}} \widehat{G}(\textbf{p},\textbf{q'})
          \left( d^{-k} \cdot Tr(\sigma_{q_1'} \otimes \cdots \otimes \sigma_{q_k'}T(\textbf{q})) \right)
  \end{align}
  By Schur-Weyl Duality, $T(\text{q})$ must be a linear combination of permutation operators ($T(\textbf{p})$ commutes with all unitaries of the form $U^{\otimes k}$ by the invariance of the Haar integral). However, by Lemma (\ref{permfix}) above, $\widehat{G}$ fixes the permuatation operators. To understand this more explicitly, rewrite the Lemmma (\ref{permfix}) as follows:
  \begin{equation}
    \sum_{\textbf{q}} \widehat{G}(\textbf{p},\textbf{q})\cdot Tr(\sigma_{q_1} \otimes \cdots \otimes \sigma_{q_k} P_\pi) = Tr(\sigma_{p_1} \otimes \cdots \otimes \sigma_{p_k}P_\pi)
  \end{equation}
  This simply expressing the equality $\widehat{G}P_\pi = P_\pi$ for permutation operator $p_\pi, \; \pi \in S_{k}$. By linearity of trace, the equality above reduces to
  \begin{align}
    \sum_{\textbf{q'}} \widehat{G}(\textbf{p},\textbf{q'})
         \left( d^{-k} \cdot Tr(\sigma_{q_1'} \otimes \cdots \otimes \sigma_{q_k'}T(\textbf{q})) \right)
         & = Tr(\sigma_{p_1} \otimes \cdots \otimes \sigma_{p_k} T(\text{q})) \\
         & = \widehat{G}(\textbf{p},\textbf{q})
  \end{align}
\end{proof}

@@

      \item $\widehat{G}(p;0) = \delta_{p0}$ (Here we denote $\widehat{G}(p;0) := \widehat{G}(p,p;0,0)$)
      \begin{proof}
        This just derives from $\widehat{G}$ acting on $I$
      \end{proof}

      \item $\widehat{G}(p,a) = \frac{1}{d^2-1}, \quad a,p \neq 0$

@@

Recall that $T(p,q)$ refers to the expected \emph{state} from applying unitary $U \sim \mathbb{U}(d)$ to the basis state $\sigma_p \otimes \sigma_q$. From the determination above, it is simple to see that $U$ uniformly samples from all diagonal basis states when acting on $\sigma_p \otimes \sigma_p, \; p \neq 0$. Strikingly enough, applying $U \sim \mathbb{U}(d)$ conserves the total sum of the coefficients of the diagonal basis states.

@@

\subsection{Convergence of the Random Circuit}
Before embarking on the Markov chain analysis, we must prove that our random circuit indeed converges to Haar distribution. This will be proven for the general case of a universal distribution (Definition (\ref{univdef})) on $\mathbb{U}(d)$. Specifcally, we will prove that any universal distribution on $\mathbb{U}(d)$ will be $k$-copy gapped (Definition (\ref{gappeddef})). In turn, this leads to convergence in sense of $\lim_{t \rightarrow \infty} (\widehat{G}_{\mu})^{t} = \widehat{G}_{\mathbb{U}(d)}$. To see this, observe that another formulation of a $k$-copy gapped distributon over $\mathbb{U}(d)$ is
\begin{equation} \label{gappeddis}
  ||G_{\mu} - G_{\mathbb{U}(d)}||_\infty < 1
\end{equation}
where $G_{\mu} = \mathbb{E}_{U \sim \mu} \left[ U^{\otimes k} \otimes (U^\dagger)^{\otimes k} \right]$. To justify this, first recall that $\widehat{G}$ and $G$ are unitarily-related by Definition \ref{Tdef}, so in proving \ref{gappeddis}, we can argue on the operators $\widehat{G}$ instead. In view of this, $\widehat{G}_{\mathbb{U}(d)}$ is a projector and thus has eigenvalues 0 and 1. If we can prove that any eigenvector of $\widehat{G}_{\mathbb{U}(d)}$ of eigenvalue one is also an eigenvector of $\widehat{G}_{\mu}$ of eigenvalue one i.e the $k$-symmetric subspace of $(\mathbb{C}^d)^{\otimes k}$ is also fixed by $\widehat{G}_{\mu}$. Thus, Equation $\ref{gappeddis}$ shows that indeed we can converge to $G_{\mathbb{U}(d)}$ by iterating our scheme (as denoted by $(\widehat{G}_\mu)^t$). Note that $\widehat{G}_{\mu}$ may not even be Hermitian, so we cannot directly argue on its spectrum (as the spectrum may not completely contain real values).
%
\begin{theorem} \label{convergence}
  Let $\mu$ be a universal distribution over $\mathbb{U}(d)$. Then all eigenvectors of $G_{\mathbb{U}(d)}$ of eigenvalue one will also be eigenvectors of $G_{\mu}$ of eigenvalue one. Furthermore, $\mu$ will be $k$-copy gapped.
\end{theorem}
%
\noindent For the sake of presentation, the proof of this will be delegated towards the appendix of this report.
%
\subsection{Markov Chain Analysis}
As discussed in a previous section, the central idea of proving \emph{rapid} convergence involves mapping the evolution of the inital state under our random circuit to that of a Markov chain's. In fact, we can explicitly determine the operator $P$ capturing our dynamics. Let $\widehat{G}^{(ij)}$ denote the unitary transformation resulting from sampling some unitary $U \sim \mathbb{U}(d)$ and applying $U$ to qubits $i$ and $j$ while leaving all other qubits undisturbed.
%
\begin{equation} \label{schemechain}
P = \frac{1}{n(n-1)}\sum_{i \neq j} \widehat{G}^{(ij)}
\end{equation}
as evidenced by the equality
\begin{equation}
  \gamma_{t+1}(\textbf{p}) = \frac{1}{n(n-1)}\sum_{i \neq j}\sum_{\textbf{q}} \widehat{G}^{(ij)}(\textbf{p},\textbf{q})\gamma_t(\textbf{q})
\end{equation}
where $\textbf{p},\textbf{q} \in \{0,1,2,3\}^2$ and $\gamma_t(\textbf{p})$ represents the expected coeffcients at timestep $t$.
%
A major caveat of interpreting $P$ as a \emph{transition matrix} for some suitable Markov chain on the diagonal states $\sigma_p \otimes \sigma_p$ is the possibility that our initial coefficients of diagonal states $\gamma_0(p,p)$ may not even sum to one or even be non-negative. It turns out that this doesn't bring too much trouble into the analysis as pointed out by Corollary \ref{maincorollary}. For now, let us tackle the case where the two conditions above are true and direct our focus on the evolution of the diagonal states. The ultimate goal of this line of thought will be to first prove Lemma \ref{mainlemma} through a carefully-done investigation of the \emph{mixing time} of $P$. Of course, we need to also show that the expected coeffcients of the non-diagonal states $\gamma_0(p,q)$ vanish rapidly as well. These matters will be addressed in a future section.

>>>>

The theory of randomness and its role in computation are cornerstones of contemporary computation theory as evidenced by decades of beautiful, celebrated results. Quantum Information has, by no means, become stranger to this trend. Numerous applications of random unitaries have been discovered in the areas of quantum tomography, quantum cryptography, and, rather naturally, quantum algorithms. To gain access to uniformly-sampled elements from $\mathbb{U}(d)$, one must sample from the Haar distribution over $\mathbb{U}(d)$ where $\mathbb{U}(d)$ denotes the group of unitary $d \times d$ matrices. One method of sampling from the Haar distribution is to apply local random unitaries sampled from distributions over lower-dimensional unitary groups. For example, if we consider an $n$-qubit system, one method of sampling from the Haar distribution over $\mathbb{U}(2^n)$ would be through the following scheme:
\begin{enumerate}
  \item Sample from the Haar distribution over $\mathbb{U}(4)$ to produce a random gate $G$
  \item Uniformly pick two distinct qubits and locally apply $G$ to the pair.
\end{enumerate}
However, this becomes computationally infeasible as even reasonably approximating the Haar distribution under this manner requires an exponential number of local 2-qubit gates \cite{knill}. This hurdle motivates us to study the construction of efficient $k$-designs i.e distributions over $\mathbb{U}(4)$ such that sampling over these distributions and applying our scheme above will converge to good approximations of the $k^{th}$ moments of the Haar distributon over $\mathbb{U}(2^n)$. To this end, we begin with annotations of the seminal paper by Harrow and Low \cite{harrlow}. We first introduce concepts germane to the area of efficient $k$-designs and summarize the techniques employed in the analyses presented in the paper.

@@

\subsection{Definitions and Motivations}
In \cite{harrlow}, the authors tackle the problem of approximate $2$-designs, specifically \emph{unitary} $2$-designs. The central result of the paper claims that the scheme of applying sampled $2$-qubit unitaries from certain distributions with a $k$-gapped property converge to the second moment of the Haar distribution rapidly. The crux of the convergence argument relies on mapping the evolution of  diagonal states to that of a Markov chain's.
Consequently, the analysis becomes amenable to classic techniques created to study mixing times of Markov chains. To begin, we first state the relevant definitions as introduced by Harrow and Low.
%
\begin{remark}
  As dissecting every proof in \cite{harrlow} may unnecessarily length this report, we occasionally omit proofs of lemmas. Some major theorems will have their proofs delegated to the appendix of this report.
\end{remark}

@@

As discussed in the introduction above, the unitary constructed by sampling from a universal gate set and applying them to random pairs of qubits will converge to a unitary sampled from Haar distribution. However, we would like to find \emph{efficient approximate} $k$-designs. Here, efficient refers to random circuits of asymptotic length polynomial in the number of qubits and $\log 1/\epsilon$. The authors actually prove some stronger, namely that \emph{$k$-copy gapped} distributions suffice.

@@

  then we say that $\mathcal{G}_\mu$ is \emph{$k$-copy gapped} if $\mathcal{G}_\mu$ has only $k!$ eigenvectors with unit eigenvalues.

@@

\noindent A more general definition for the universality of gatesets can be defined as well:

@@

  Let $\mu$ be a distribution over $\mathbb{U}(d)$. Then $\mu$ is deemed to be \emph{universal} for $\mathbb{U}(d)$ if for any open ball $B \subset \mathbb{U}(d)$, there exists a sufficiently large integer $\ell$ such that $\mu^{\star \ell}(B) > 0$

@@

  This is the convolution product in the context of \emph{measures}. For example, if we let $\mu,\sigma$ be two finite Borel measures over $\mathbb{U}(d)$, then we have a new finite Borel measure on $\mathbb{U}(d)$, $\mu^{\star \ell}$:
  $$ \mu^{\star \ell} (S) = \int_{\mathbb{U}(d)}\cdots\int_{\mathbb{U}(d)} \mathbb{I}_S(U_1\cdots U_{\ell}) d\mu(U_1) \cdots d\mu(U_\ell) $$
  where $\mathbb{I}_S$ is the characteristic function for measurable set $S \subset \mathbb{U}(d)$.

@@

\noindent Intuitively, universality implies that sampling enough times over $\mu$ is enough to approximate any gate in $\mathbb{U}(d)$. This aligns with the definition introduced for discrete measures $\mu$ over finite gate sets. We will prove later that any universal gate set will be $k$-copy gapped on $\mathbb{U}(4)$ (Theorem  \ref{convergence}). For now, let us state the main theorem and its auxiliary lemma to guide our upcoming analysis:

@@

  Let $\rho$ be an initial state of an $2n$-qubit system. We can expand $\rho$ in terms of the Pauli basis $\{\sigma_i\}_{0 \leq i \leq 3}$ as follows:

  \begin{equation}
    \rho = \frac{1}{2^n} \sum_{p_1,p_2} \gamma_0(p_1,p_2) \sigma_{p_1} \otimes \sigma_{p_2}, \quad \gamma_0(p_1,p_2) \in \mathbb{C}
   \end{equation}
   %
  where $p_1,p_2$ run over all strings of Pauli indices of length $n$, $\{0,1,2,3\}^n$. Let $\gamma_W(p_1,p_2)$ be the coefficient of the term $\sigma_{p_1} \otimes \sigma_{p_2}$ for the output state after applying random circuit $W$, $\rho_W = W^{\otimes 2} \rho (W^\dagger)^{\otimes 2}$. Extending this notation allows us to express the expected coefficient after $t$-samples from our distribution $\mu$ as
  $\gamma_t(p_1,p_2) = \mathbb{E}_W [\gamma_W(p_1,p_2)]$. Note that the expectation is taken over all $t$-length random circuits.
\end{notation}
%
\begin{lemma} \label{mainlemma}
  Let $\rho$ be a an initial state such that $\gamma_0(p,p) \geq 0$ and $\sum_{p} \gamma_0(p,p) = 1$ with $\mu$ and $W$ as in Theorem \ref{maintheorem}. Then there exists constant $C$ such that for $\epsilon > 0$

@@

  \begin{equation}
    \sum_{p_1,p_2 \neq 00} \left| \gamma_t(p_1,p_2) - \delta_{p_1p_2} \frac{1}{2^n(2^n + 1)} \right| \leq \epsilon
  \end{equation}
  for $t \geq Cn(n + \log 1/\epsilon)$
\end{lemma}

\begin{corollary} \label{maincorollary}
  Let $\mu$, $W$, $\gamma_W$ be as above. Then for \emph{\bf any} initial state $\rho$, there exists constant $C$ such that for $\epsilon > 0$,
  \begin{align}
    & \sum_{p_1,p_2 \neq 00} \left( \gamma_t(p_1,p_2) - \delta_{p_1p_2}\frac{\sum_{p\neq 0} \gamma_0(p,p)}{4^n - 1} \right)^2 \leq \epsilon \\
    & \sum_{p_1,p_2 \neq 00} \left| \gamma_t(p_1,p_2) - \delta_{p_1p_2}\frac{\sum_{p\neq 0} \gamma_0(p,p)}{4^n - 1} \right| \leq \epsilon
  \end{align}
  both for $t \geq Cn(n + \log 1/ \epsilon)$
\end{corollary}

@@

We turn back to the general case and begin to define concepts and definitions for group $\mathbb{U}(d)$ and for $d$-qudit states. Set a Hermitian basis of $d \times d$ to be $\sigma_0,\sigma_1,\cdots \sigma_{d^2 -1}$ such that $\sigma_0 = I$. These basis elements span the $d$-qudit states. Here we stress that we sample our unitaries over the \emph{Haar distribution} over $\mathbb{U}(d)$. We denote sampling from the Haar distribution with the notation $U \sim \mathbb{U}(d)$ when necessary. Furthermore, our Hermitian basis operators are such that $Tr(\sigma_p\sigma_q) = d \cdot \delta_{p,q}$ i.e orthogonal under the Hilbert-Schmidt inner product. \newline

\noindent Let ${\bf p} = (p_1, \cdots, p_k) \in \{0,1,\cdots, d^2 -1\}^k$ denote a $k$-string of basis indices. We will be interested in the quantity:
%
\begin{equation} \label{Tdef}
  T({\bf p}) := \mathbb{E}_{U \sim \mathbb{U}(d)} \left[ U^{\otimes k} \sigma_{p_1} \otimes \sigma_{p_2} \otimes \cdots \otimes \sigma_{p_k} (U^\dagger)^{\otimes k} \right]
\end{equation}
This will be the expected state after applying a sampled unitary $U$ from the Haar distribution over $\mathbb{U}(d)$. Naturally, we can reexpress this resulting state into the $\sigma_p$ basis:
%
\begin{equation}
  T({\bf p}) = \sum_{{\bf q}} \widehat{G}(\textbf{p},\textbf{q}) \sigma_{q_1} \otimes \cdots \otimes \sigma_{q_k}
\end{equation}
where the same is taken over all $k$-strings in $\{0, \cdots, d^2-1\}^k$. This allows us to define the matrix $\widehat{G}$ where the rows and columns are indiced by $k$-strings with entries $\widehat{G}(\textbf{p},\textbf{q})$. By virtue of the invariance of the Haar distribution, this matrix turns out to be a \emph{symmetric projector}.

\begin{lemma}
  The matrix $\widehat{G}$ above is symmetric i.e. $\widehat{G}(\textbf{p},\textbf{q}) = \widehat{G}(\textbf{q},\textbf{p})$
\end{lemma}
%
\begin{proof}
  By orthogonality:
  \begin{align}
    \widehat{G}(\textbf{q},\textbf{p})
    & = d^{-k} \cdot Tr(\sigma_{q_1} \otimes \cdots \otimes \sigma_{q_k} T(\textbf{p}) ) \\
    & = d^{-k} \mathbb{E}_{U}\left[ Tr(\sigma_{q_1} \otimes \cdots \otimes \sigma_{q_k}) U^{\otimes k} (\sigma_{p_1} \otimes \sigma_{p_2} \otimes \cdots \otimes \sigma_{p_k}) (U^\dagger)^{\otimes k}\right]
  \end{align}
\end{proof}
%
\noindent Note that the expectation and the trace operator commute by linearity of latter. The result follows from the cyclic invariance of trace.
%
\begin{lemma} \label{permfix}
$\widehat{G}$ fixes the permutation operators $P_{\pi}, \; \pi \in S_k$.
\end{lemma}
%
\noindent These two results are all that is required to show that $\widehat{G}$ is indeed a projector.
%
\begin{theorem} \label{g-proj}
  $\widehat{G}$ is a projector i.e. $\widehat{G}^2 = \widehat{G}$
\end{theorem}
%
\begin{proof}
  Expanding the matrix product shows that
  \begin{align}
     \widehat{G}(\textbf{p},\textbf{q})
     & = \sum_{\textbf{q'}} \widehat{G}(\textbf{p},\textbf{q'})\widehat{G}(\textbf{q'},\textbf{q}) = \\
     & = \sum_{\textbf{q'}} \widehat{G}(\textbf{p},\textbf{q'})
          \left( d^{-k} \cdot Tr(\sigma_{q_1'} \otimes \cdots \otimes \sigma_{q_k'}T(\textbf{q})) \right)
  \end{align}
  By Schur-Weyl Duality, $T(\text{q})$ must be a linear combination of permutation operators ($T(\textbf{p})$ commutes with all unitaries of the form $U^{\otimes k}$ by the invariance of the Haar integral). However, by Lemma (\ref{permfix}) above, $\widehat{G}$ fixes the permuatation operators. To understand this more explicitly, rewrite the Lemmma (\ref{permfix}) as follows:
  \begin{equation}
    \sum_{\textbf{q}} \widehat{G}(\textbf{p},\textbf{q})\cdot Tr(\sigma_{q_1} \otimes \cdots \otimes \sigma_{q_k} P_\pi) = Tr(\sigma_{p_1} \otimes \cdots \otimes \sigma_{p_k}P_\pi)
  \end{equation}
  This simply expressing the equality $\widehat{G}P_\pi = P_\pi$ for permutation operator $p_\pi, \; \pi \in S_{k}$. By linearity of trace, the equality above reduces to
  \begin{align}
    \sum_{\textbf{q'}} \widehat{G}(\textbf{p},\textbf{q'})
         \left( d^{-k} \cdot Tr(\sigma_{q_1'} \otimes \cdots \otimes \sigma_{q_k'}T(\textbf{q})) \right)
         & = Tr(\sigma_{p_1} \otimes \cdots \otimes \sigma_{p_k} T(\text{q})) \\
         & = \widehat{G}(\textbf{p},\textbf{q})
  \end{align}
\end{proof}

@@

      \item $\widehat{G}(\textbf{p};\textbf{0}) = \delta_{p0}$ (Here we denote $\widehat{G}(\textbf{p};\textbf{0}) := \widehat{G}(p,p;0,0)$. )
      \begin{proof}
        This just derives from $\widehat{G}$ acting on $I$
      \end{proof}

      \item $\widehat{G}(\textbf{p},\textbf{a}) = \frac{1}{d^2-1}, \quad \textbf{p},\textbf{a} \neq 0$

@@

Recall that $T(p,q)$ refers to the expected state from applying unitary $U \sim \mathbb{U}(d)$ to the basis state $\sigma_p \otimes \sigma_q$. From the determination above, it is simple to see that $U$ uniformly samples from all diagonal basis states when acting on $\sigma_p \otimes \sigma_p, \; p \neq 0$. Strikingly enough, applying $U \sim \mathbb{U}(d)$ conserves the total sum of the coefficients of the diagonal basis states.

@@

\subsection{Convergence of the Random Circuit}
Before embarking towards Markov chain techniques, we must prove that our random circuit construction indeed converges to Haar distribution. Specifcally, we will prove that any universal distribution (Definition \ref{univdef}) on $\mathbb{U}(d)$ will be $k$-copy gapped (Definition \ref{gappeddef}). In turn, this leads to convergence in sense of $\lim_{t \rightarrow \infty} (\widehat{G}_{\mu})^{t} = \widehat{G}_{\mathbb{U}(d)}$. To see this, observe that another formulation of a $k$-copy gapped distributon over $\mathbb{U}(d)$ is
\begin{equation} \label{gappeddis}
  ||G_{\mu} - G_{\mathbb{U}(d)}||_\infty < 1
where $G_{\mu} = \mathbb{E}_{U \sim \mu} \left[ U^{\otimes k} \otimes (U^\dagger)^{\otimes k} \right]$. To justify this, first recall that $\widehat{G}$ and $G$ are unitarily-related by Definition \ref{Tdef}, so in proving \ref{gappeddis}, we can argue on the operators $\widehat{G}$ instead. In view of this, $\widehat{G}_{\mathbb{U}(d)}$ is a projector and thus has eigenvalues 0 and 1. If we can prove that any eigenvector of $\widehat{G}_{\mathbb{U}(d)}$ of eigenvalue one is also an eigenvector of $\widehat{G}_{\mu}$ of eigenvalue one i.e the $k$-symmetric subspace of $(\mathbb{C}^d)^{\otimes k}$ is also fixed by $\widehat{G}_{\mu}$. Thus, Equation $\ref{gappeddis}$ shows that indeed we can converge to $G_{\mathbb{U}(d)}$ as all eigenvectors shrink by iterating our construction. Note that $\widehat{G}_{\mu}$ may not even be Hermitian, so we cannot directly argue on its spectrum (as the spectrum may not completely contain real eigenvalues).
\begin{theorem} \label{convergence}
  Let $\mu$ be a universal distribution over $\mathbb{U}(d)$. Then all eigenvectors of $G_{\mathbb{U}(d)}$ of eigenvalue one will also be eigenvectors of $G_{\mu}$ of eigenvalue one. Furthermore, $\mu$ will be $k$-copy gapped.
\noindent For the sake of presentation, the proof of this will be delegated towards the appendix of this report.
\subsection{Markov Chain Analysis}
As discussed in a previous section, the central idea of proving \emph{rapid} convergence involves mapping the evolution of the diagonal states under our random circuit to that of a Markov chain's. Here, we refer to states of the form
$$ \psi = \frac{1}{2^n}\sum_{p}\gamma(p,p) \sigma_{p} \otimes \sigma_{p}, \quad p \in \{0,1,2,3\}^n, \; \gamma(p,p) \in \mathbb{C} $$
In fact, we can explicitly determine the operator $P$ capturing our dynamics. Let $\widehat{G}^{(ij)}$ denote the unitary transformation resulting from sampling some unitary $U \sim \mathbb{U}(4)$ and applying $U$ to qubits $i$ and $j$ while leaving all other qubits undisturbed.
\begin{equation} \label{schemechain}
P = \frac{1}{n(n-1)}\sum_{i \neq j} \widehat{G}^{(ij)}
as evidenced by the equality
  \gamma_{t+1}(\textbf{p}) = \frac{1}{n(n-1)}\sum_{i \neq j}\sum_{\textbf{q}} \widehat{G}^{(ij)}(\textbf{p},\textbf{q})\gamma_t(\textbf{q})
where $\textbf{p},\textbf{q} \in \{0,1,2,3\}^2$ and $\gamma_t(\textbf{p})$ represents the expected coeffcients at timestep $t$.
A major caveat of interpreting $P$ as a \emph{transition matrix} for some suitable Markov chain on the diagonal states is the possibility that our initial coefficients of the diagonal states $\gamma_0(p,p)$ may not even sum to one or even be non-negative. It turns out that this doesn't bring too much trouble into the analysis as pointed out by Corollary \ref{maincorollary}. For now, let us tackle the case where the two conditions above are true and direct our focus on the evolution of the diagonal states. The ultimate goal of this line of thought will be to first prove Lemma \ref{mainlemma} through a carefully-done investigation of the \emph{mixing time} of $P$. Of course, we need to also show that the expected coeffcients of the non-diagonal states $\gamma_0(p,q)$ vanish rapidly as well. These matters will be addressed in a future section.
